---
title: "Biostat 203B Homework 1 Solutions"
author: Analissa Avila
subtitle: Due Jan 22 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=F}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

    **Solution:** Done
  
2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

    **Solution:** Done
  
3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

    **Solution:** Done
  
4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

    **Solution:** Done
  
5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

    **Solution:** Done
  
  
## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)


    **Solution:**  
    Here is the link to my [course completion report](https://www.citiprogram.org/verify/?ke701e4f3-cd9e-4b2c-bb4d-9913085b5c57-40340162).  
    Here is the link to my [course completion certificate](https://www.citiprogram.org/verify/?w30eef218-8d98-4cf0-836b-9b0c350f5427-40340162).  
  
  
2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  
    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

    Use Bash commands to answer following questions.  
    
    **Solution:** Done
  
  
3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

    **Solution:**  
    ```{bash}
    ## contents in 'core' folder
    ls -l /usr/203b-data/mimic-iv/core
    ```
    
    ```{bash}
    ## contents in 'hosp' folder
    ls -l /usr/203b-data/mimic-iv/hosp
    ```
    
    ```{bash}
    ## contnets in 'icu' folder
    ls -l /usr/203b-data/mimic-iv/icu
    
    ```
    
    The bash commands `zcat`, `zless`, `zmore`, and `zgrep` are equivalent to `cat`, `less`, `more` and `grep`, but they operate on the uncompressed contents of compressed files.  
    
    `zcat` prints the contents of the compressed file   
    `zless` browses a compressed text file screen by screen (upwards or downwards)    
    `zmore`browses a compressed text file screen by screen (only downwards)    
    `zgrep`prints lines of a compressed file that match an expression  
      
      
      
4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

    **Solution:**  
    This script prints detailed information on all compressed files (.gz) inside the `core` folder
    ```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```

    This script prints the number of lines in each data file in the `core` folder
    ```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        echo "$(zcat $datafile | wc -l) $datafile"
      done
    ```
  
  
5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

    **Solution:**  
    First few lines of `admissions.csv.gz`
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head
    ```
    
    Number of rows in `admissions.csv.gz`
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | wc -l
    ```
    
    Number of unique patients in `admissions.csv.gz`
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n +2 | \
    awk -F ',' '{print $1}' | sort | uniq | wc -l
    ```
    
    Possible values taken by `admission_type` and their counts
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n +2 | \
    awk -F ',' '{print $6}' | sort | uniq -c
    ```
    
    Possible values taken by `admission_location` and their counts
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n +2 | \
    awk -F ',' '{print $7}' | sort | uniq -c
    ```
    
    Possible values taken by `insurance` and their counts
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n +2 | \
    awk -F ',' '{print $9}' | sort | uniq -c
    ```
    
    Possible values taken by `language` and their counts
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n +2 | \
    awk -F ',' '{print $10}' | sort | uniq -c
    ```
    
    Possible values taken by `marital_status` and their counts
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n +2 | \
    awk -F ',' '{print $11}' | sort | uniq -c
    ```
    
    Possible values taken by `ethnicity` and their counts
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n +2 | \
    awk -F ',' '{print $12}' | sort | uniq -c
    ```
    


## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

    **Solution:** Elizabeth is the most mentioned character.
    ```{bash}
    for name in Elizabeth Jane Lydia Darcy
     do
        echo "${name} $(grep -o ${name} pride_and_prejudice.txt | wc -l)"
     done
    ```


2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```
    
    **Solution:**  
    The first command uses `>` which directs the output "hello, world" to a file named test1.txt. If that command is run again, a new file named test1.txt will be created.  
    The second command uses `>>` which appends the output "hello, world" to a file named test2.txt. If that command is run again, another line of "hello, world" will be appended to the existing test2.txt file.  


3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

    **Solution:**
    ```{bash eval=FALSE}
    #Run this command in local machine to copy file to teaching server
    scp middle.sh avila.analissa@server.ucla-biostat-203b.com:/home/avila.analissa/biostat-203b-2021-winter/hw1/
    
    ```
    
    ```{bash}
    ## make file executable by the owner 
    chmod u+x middle.sh
    
    ## run
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
    
    The output is the last 5 lines of the first 20 lines (lines 16-20) of the Pride and Prejudice text file. Specifically, it prints the author and editor information.  
    The `"$1"`, `"$2"`, and `"$3"` are positional parameters. They contain the content of the command line:`$1` contains the first item (pride_and_prejudice.txt - file name), `$2` contains the second item (20 - # of first lines to print using `head` command), and `$3` contains the third item (5 - # of last lines to print using `tail` command).  
    The first line tells Linux to use the 'sh' shell to run the commands 


## Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution:**

`cal` displays the current month's calendar
```{bash}
cal
```

`cal 2021` displays the full 2021 calendar 
```{bash}
cal 2021
```

`cal 9 1752` displays the September 1752 calendar. The unusual part of this calendar is that the days September 3-13 are skipped. This is because in 1752 Britain adopted the Gregorian calendar. 
```{bash}
cal 9 1752
```

`date` displays the current date and time in UTC
```{bash}
date
```

`hostname` displays the name of the current system
```{bash}
hostname
```

`arch` displays the machine architecture for the current machine
```{bash}
arch
```

`uname -a` displays the following system information: kernel name, node name, kernel release, kernel version, machine hardware name, processor type, hardware platform, and operating system
```{bash}
uname -a
```

`uptime` displays how long the system has been running, the current time, the number users currently logged-in, and the average system load for the last 1, 5, and 15 minutes
```{bash}
uptime
```

`who am i` returns nothing, but if you omit the spaces `whoami` displays the current user's user name
```{bash}
who am i

whoami
```

`who` displays a information about each user who is currently logged-in. The information includes: user name, terminal name, login time, and remote host  
```{bash}
who
```

`w` displays the `uptime` information as a header, along with more/better formatted information about who is currently logged-in. The current user information includes: user name, terminal name, remote host, login time, idle time, JCPU time, PCPU time, and command line of their current process
```{bash}
w
```

`id` displays the real and effective user and group id's for the current user
```{bash}
id
```

`last | head` displays the first 10 lines of the list of all users who have logged in and out of the system (starting with the most recent user to login). The list includes the user's name, terminal, remote host, and login and logout time.
```{bash}
last | head
```

`echo {con,pre}{sent,fer}{s,ed}` prints all 8 words that are created by sequentially joining one element from each bracketed set. For example, combining the first element of each bracketed set gives the word 'consents'
```{bash}
echo {con,pre}{sent,fer}{s,ed}
```

`time sleep 5` this command tells the system to pause for 5 seconds and then print the time it took for the command to run (which should be very close to 5 seconds)
```{bash}
time sleep 5
```

`history | tail` this command should display the last 10 lines of the user's command history (note you have to enable history or the command will return nothing). Since `history | tail` is the only command in this chunck, it is the only command returned
```{bash}
## first enable history
set -o history

history | tail
```